---
title: "Class 12 Lab"
author: "Derek Chang (PID: A16942232)"
format: pdf
---

## Population Scale:
One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

How many samples do we have: 

Reading the Table: 
```{r}
expr <- read.table("DataClass12.txt")
head(expr)

```

```{r}
summary(expr)
```

Reading the Sample size for each genotype:

```{r}
table(expr$geno)
```


# Making a boxplot from expr data
```{r}
library(ggplot2)
```

Making the boxplot using `ggplot`:

```{r}
exprplot <- ggplot(expr, aes(geno, exp, fill = geno)) + geom_boxplot()
exprplot
```


>Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes

```{r}
library(dplyr)
medians <- expr %>% group_by(geno) %>% summarize(median_value = median(exp))
medians
```


Given the table function for each genotype, We have 108 A/A,  233 A/G, and 121 G/G genotypes. The median for A/A is 31.24847, the median for A/G is 25.06486, and the median for G/G is 20.07363.



>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Based on the boxplot, we can infer that A/A has a much higher relative expression value compared to G/G. Given this, it seems that the SNP does affect the expression levels of ORMDL3.





